Equalizer 2.0 --\\Convergence of a Parallel Rendering Framework

### 2018-02-08 paperIEEE.tex
Rework & results
git-svn-id: https://dev.ifi.uzh.ch/vmml/svn/Equalizer16@108 db7c848b-7755-4df2-b763-8796797afa52

In particular, we present novel algorithms for parallel rendering which did not yet
appear in a separate publication, including pixel and sub-pixel decompositions,
dynamic frame resolution, tunable sort-first load-balancing parameters, frame-rate
equalization, thread synchronization modes for multi-threaded rendering, a
powerful abstraction for multi-view rendering on arbitrary display setups,
dynamic focus distance and asymmetric eye positions for VR,
parallel pixel streaming to tiled display walls, as well as a fully-fledged data
distribution API with compression and reliable multicast.

@@

On the other hand, software for driving and interacting with tiled display walls
has received significant attention, including \textsf{Sage}~\cite{Sage} and
\textsf{Sage~2}~\cite{Sage2} in particular. \textsf{Sage} was built entirely
around the concept of a shared framebuffer where all content windows are
separate applications using pixel streaming. It is no longer actively supported.
\textsf{Sage 2} is a complete, browser-centric reimplementation where each
application is a web application distributed across browser instances.
\textsf{DisplayCluster}~\cite{DC}, and its continuation
\textsf{Tide}~\cite{tide}, also implement the shared framebuffer concept of
\textsf{Sage}, but provide a few native content applications integrated into the
display servers. All these solutions implement a scalable display environment and
are a target display platform for scalable 3D graphics applications.

@@

Furthermore, various applications and frameworks have used \textsf{Equalizer}
for new research in visualization. On the application side, \textsf{RTT
Deltagen}, \textsf{Bino}, \textsf{Livre} and \textsf{RTNeuron}~\cite{HBBES:13}
are the most mature examples and are presented in \sref{sApplications}. On the
framework side, \textsf{Omegalib}~\cite{Omegalib}, a framework used in the
Cave2, made significant progress in integrating 2D collaborative workspaces like
\textsf{Sage~2} with 3D immersive content. Lambers et.al. developed a framework
for visualizing remote sensing data~\cite{LK:09} on large displays and immersive
installations based on Equalizer.

@@

This section introduces new decomposition modes, runtime adaptations and other
optimizations to increase the rendering performance for a wide variety of
algorithms and use cases.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
following we present the newly added decomposition modes and motivate their use
case, which bring the overall feature set way beyond the typical sort-first and
sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
modes. The way how compounds are setup is still as given in \cite{EMP:09}.

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks, or work packages are queued and
processed by all source channels by polling a server-central queue. Prefetching
ensures that the task communication overlaps with rendering. As shown in~
\cite{SPEP:16}, these modes can provide better performance due to being inherently
load-balanced, as long as there is an insignificant overhead for the render task
setup. This mode is transparent to \textsf{Equalizer} applications.

@@

Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
decompose the work for a single pixel, for example with Monte-Carlo ray tracing,
FSAA or depth of field rendering. Composition typically uses accumulation and
averaging of all computed fragments for a pixel. This feature is not fully
transparent to the application, since it needs to adapt (jitter or tilt) the
frustum based on the iteration executed. Furthermore, subpixel compounds
interact with idle image refinements, e.g., they can accelerate idle
anti-aliasing of a scene when the camera and scene are not changed.

@@

We have implemented various tunable parameters for both equalizers allowing
application developers to optimize the load balancing based on the
characteristics of their rendering algorithm:

@@

The monitor equalizer (\fig{fig:monitor}) allows to reuse the rendering on
another channel, typically for monitoring a larger setup on a control
workstation. Output frames on the display channels are connected to input frames
on a single monitoring channel. The monitor equalizer changes the scaling factor
and offset between the output and input, so that the monitor channel has the
same, but typically downscaled view, as the originating segments.

@@

The region of interest is the screen-space 2D bounding box enclosing the
geometry rendered by a single resource. We have extended the core parallel
rendering framework to use an application-provided ROI to optimize the load
equalizer as well as image compositing performance. The load equalizer uses the
ROI to refine its load grid to the regions containing data. The compositing code
uses the ROI to minimize image readback and network transmission. In
\cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of the algorithm, and
show that using ROI can quadruple the rendering performance, in particular for
the costly compositing step in sort-last rendering.

@@

Compression for the compositing step in parallel rendering is critical for
performance. This not only applies to the well-researched network transfer step,
but also for the transfer between GPU and CPU. \textsf{Equalizer} supports a
variety of
compression algorithms, from very fast RLE encoding, YUV
subsampling on the GPU to JPEG compression. These algorithms are implemented as
runtime-loaded
plugins, allowing easy extension and customization to application-specific
compression. In \cite{MEP:10}, we show this to be a critical step for
interactive performance at scale.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
many complex applications and large data scenarios. To easily customize the
synchronization of all threads on a single process, \textsf{Equalizer}
implements three threading modes:  Full synchronization, draw synchronization
and asynchronous. Note that the execution between node processes is always
asynchronous, for up to \textsf{latency} frames.

@@

In draw synchronization, the node thread and all render threads are synchronized
for all \textsf{frameDraw} operations; that is, \textsf{Node::frameFinish} is
executed after the last channel is done drawing. This allows the render threads
to read shared data during their draw operation, but not during compositing.
Since compositing is typically independent of the rendered data, this is the
default mode. This mode allows to overlap compositing with rendering and data
synchronization on multi-GPU machines.

In asynchronous execution, all threads run asynchronously. Render threads may
work on different frames at any given time. This mode is the fastest, and
requires the application to have one instance of each mutable object in each
render thread. It is required for using time-multiplex compounds in
multithreaded rendering.

@@

Head tracking is the minimal feature needed to support immersive installations.
\textsf{Equalizer} does support multiple, independent tracked views through the
observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
direct, application-transparent configuration of a VRPN tracker device.
Alternatively, an application can provide a $4\times 4$ tracking matrix defining
the transformation from the canvas coordinate system to the observer. Both
CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
displays are implemented.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current view direction of the user.
Initial experiments show that this provides better viewing comfort, in
particular for objects placed in front of the physical displays.

@@

Traditional head tracking computes the left and right eye positions by using a
configurable interocular distance. However, human heads are not symmetric, and
by measuring individual users a more precise frustum can be computed.
\textsf{Equalizer} supports this through the optional configuration (file or
programmatically) of individual 3D translations for all eyes relative to the
tracking matrix.

@@

\textsf{Tide} (Tiled interactive display environment) is an improved version of
\textsf{DisplayCluster}~\cite{DC}, providing a touch-based, multi-window user
interface for high-resolution tiled display walls. Remote applications receive
input events and send pixel streams using the \textsf{Deflect} client library.
\textsf{Equalizer} includes full support, enabling application-transparent
integration with \textsf{Tide}. When a \textsf{Tide} server is configured, all
output channels of a view stream in parallel to one window on the wall. In
\cite{deflect}, we have shown interactive framerates for a 24 megapixel
resolution over a WAN link. Events received are translated into
\textsf{Equalizer} events and injected into the normal event flow, allowing
seamless application integration.

@@

We investigated the scalability of individual decomposition modes by running the same experiment using a varying number of render nodes (2-9), as well as one dedicated application/display node. We subsequently summed up the duration of all rendered frames for each run (\fig{fCompounds}).

@@

This figure gives an impression of how individual modes perform on heterogeneous
systems. In this case the tree equalizer performs best (\fig{fEqualizers}
(left)), as it allows us to a priori define how much {\em usage} it should make
of individual nodes, i.e., bias the allocation of rendering time in accordance
with the (simulated) compute resources. \fig{fEqualizers} (right), on the other
hand, illustrates how the tested decomposition modes perform on a system where
compute resources fluctuate randomly every frame, as can arguably be the case
for, e.g., cloud-based applications. For this scenario tile equalizer seems best
suited, as it performs load balancing implicitly and does not assume coherence
of available resources between frames.

@@

\begin{IEEEbiography}[{
\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
Stefan Eilemann} Stefan was the technical manager of the Visualization Team in
the Blue Brain Project. His work involves working towards large-scale
visualization for Exascale simulations; interactive integration of simulation,
analysis and visualization as well as flexible frameworks for data sharing and
dynamic allocation of heterogenous resources. He is the CEO and founder of
Eyescale and also the lead developer of the \textsf{Equalizer} parallel
rendering framework. He received his masters degree in Computer Science from
EPFL in 2015, and his Engineering Diploma in Computer Science in 1998. He is
currently working towards a PhD in Computer Science at the Visualization and
MultiMedia Lab at the University of Zurich. \end{IEEEbiography}

>>>>

We present novel algorithms for parallel rendering which did not appear in a
separate publication, including pixel and sub-pixel decompositions, dynamic
frame resolution, tunable sort-first load-balancing parameters, frame-rate
equalization, thread synchronization modes for multi-GPU rendering, a powerful
abstraction for multi-view rendering on arbitrary display setups, dynamic focus
distance and asymmetric eye positions for VR, parallel pixel streaming to tiled
display walls, as well as a fully-fledged data distribution API with compression
and reliable multicast.

@@

On the other hand, software for driving and interacting with tiled display walls
has received significant attention, including \textsf{Sage}~\cite{Sage} and
\textsf{Sage~2}~\cite{Sage2} in particular. \textsf{Sage} was built entirely
around the concept of a shared framebuffer where all content windows are
separate applications using pixel streaming. It is no longer actively supported.
\textsf{Sage~2} is a complete, browser-centric reimplementation where each
application is a web application distributed across browser instances.
\textsf{DisplayCluster}~\cite{DC}, and its continuation
\textsf{Tide}~\cite{tide}, also implement the shared framebuffer concept of
\textsf{Sage}, but provide a few native content applications integrated into the
display servers. All these solutions implement a scalable display environment and
are a target display platform for scalable 3D graphics applications.

@@

Furthermore, various applications and frameworks have used \textsf{Equalizer}
for new research in visualization. On the application side, \textsf{RTT
Deltagen}, \textsf{Bino}, \textsf{Livre} and \textsf{RTNeuron}~\cite{HBBES:13}
are the most mature examples and are presented in \sref{sApplications}. On the
framework side, \textsf{Omegalib}~\cite{Omegalib}, a framework used in the
Cave2, made significant progress in integrating 2D collaborative workspaces like
\textsf{Sage~2} with 3D immersive content. Lambers et.al. developed a framework
for visualizing remote sensing data~\cite{LK:09} on large displays and immersive
installations.

@@

% This section introduces new decomposition modes, runtime adaptations and other
% optimizations to increase the rendering performance for a wide variety of
% algorithms and use cases.

@@

The initial version of \textsf{Equalizer} implemented the basic sort-first (2D),
sort-last (DB), stereo (EYE) and multilevel decompositions \cite{EMP:09}. In the
following we present the newly added decomposition modes and motivate their use
case, which bring the overall feature set way beyond the typical sort-first and
sort-last rendering modes. \fig{fig:compounds} provides an overview of the new
modes. The compound concept to set up scalable rendering is presented in
\cite{EMP:09}.

@@

Tile (\fig{fig:tiles}) and chunk decompositions are a variant of sort-first and
sort-last rendering, respectively. They decompose the scene into a predefined
set of fixed-size image tiles or database ranges. These tasks, or work packages,
are queued and processed by all source channels by polling a server-central
queue. Prefetching ensures that the task communication overlaps with rendering.
As shown in~ \cite{SPEP:16}, these modes can provide better performance due to
being inherently load-balanced, as long as there is an insignificant overhead
for the render task setup. This mode is transparent to \textsf{Equalizer}
applications.

@@

Subpixel compounds (\fig{fig:subpixel}) are similar to pixel compounds, but they
decompose the work for a single pixel, for example with Monte-Carlo ray tracing,
FSAA or depth of field rendering. Composition typically uses accumulation and
averaging of all computed fragments for a pixel. This feature is not fully
transparent to the application, since it needs to adapt (jitter or tilt) the
frustum based on the iteration executed.

% Furthermore, subpixel compounds interact with idle image refinements, e.g., they
% can accelerate idle anti-aliasing of a scene when the camera and scene are not
% changed.

@@

Both equalizers implement tuneable parameters allowing application developers to
optimize the load balancing based on the characteristics of their rendering
algorithm:

@@

The monitor equalizer (\fig{fig:monitor}, \fig{fLayout}) allows to reuse the
rendering on another channel, typically for monitoring a larger setup on a
control workstation. Output frames on the display channels are connected to
input frames on a single monitoring channel. The monitor equalizer changes the
scaling factor and offset between the output and input, so that the monitor
channel has the same, but typically downscaled view, as the originating
segments.

@@

The region of interest is the screen-space 2D bounding box enclosing the data
rendered by a single resource. We have extended the core parallel rendering
framework to use an application-provided ROI to optimize the
\texsf{load\_equalizer} as well as the image compositing. The load equalizer
uses the ROI to refine its load grid to the regions containing data. The
compositing code uses the ROI to minimize image readback and network
transmission. In \cite{MEP:10} and \cite{EBAHMP:12}, we provide the details of
the algorithm, and show that using ROI can quadruple the rendering performance,
in particular for the costly compositing step in sort-last rendering.

@@

Compression for the compositing step is critical for performance. This not only
applies to the well-researched network transfer step, but also for the transfer
between GPU and CPU. \textsf{Equalizer} supports a variety of compression
algorithms, from very fast RLE encoding, YUV subsampling on the GPU to JPEG
compression. These algorithms are implemented as runtime-loaded plugins,
allowing easy extension and customization to application-specific compression.
In \cite{MEP:10}, we show this to be a critical step for interactive performance
at scale.

@@

Different applications have different degrees on how decoupled and thread-safe
the rendering code is from the application logic. For full decoupling all
mutable data has to have a copy in each render thread, which is not feasible in
most applications and large data scenarios. To easily customize the
synchronization of all threads on a single process, \textsf{Equalizer}
implements three threading modes:  Full synchronization, draw synchronization
and asynchronous. Note that the execution between node processes is always
asynchronous, for up to \textsf{latency} frames.

@@

In draw synchronization, the node thread and all render threads are synchronized
for all \textsf{frameDraw} operations; that is, \textsf{Node::frameFinish} is
executed after the last channel is done drawing. This allows the render threads
to read shared data during their draw operation, but not during compositing.
Since compositing is often independent of the rendered data, this is the default
mode. This mode allows to overlap compositing with rendering and data
synchronization on multi-GPU machines.

In asynchronous execution, all threads run asynchronously. Render threads may
work on different frames at any given time. This mode is the fastest, and
requires the application to have one instance of each mutable object in each
render thread. It is required for scaling time-multiplex compounds on multi-GPU
machines.

@@

Head tracking is the minimal feature needed to support immersive installations.
\textsf{Equalizer} does support multiple, independent tracked views through the
observer abstraction (\sref{sec:observer}). Built-in VRPN support enables the
direct, application-transparent configuration of a VRPN tracker device.
Alternatively, applications can provide a $4\times 4$ tracking matrix. Both
CAVE-like tracking with fixed projection surfaces and HMD tracking with moving
displays are implemented.

@@

To our knowledge, all parallel rendering systems have the focal plane coincide
with the physical display surface. For better viewing comfort, we introduce a
new dynamic focus mode, where the application defines the distance of the focal
plane from the observer, based on the current \textit{lookat} distance. Initial
experiments show that this provides better viewing comfort, in particular for
objects placed in front of the physical displays.

@@

Traditional head tracking computes the left and right eye positions by using an
interocular distance. However, human heads are not symmetric, and by measuring
individual users a more precise frustum can be computed. \textsf{Equalizer}
supports this through the optional configuration of individual 3D eye
translations relative to the tracking matrix.

@@

\textsf{Tide} (Tiled interactive display environment) is an improved version of
\textsf{DisplayCluster}~\cite{DC}, providing a touch-based, multi-window user
interface for high-resolution tiled display walls. Remote applications receive
input events and send pixel streams using the \textsf{Deflect} client library.
\textsf{Equalizer} includes full support, enabling application-transparent
integration with \textsf{Tide}. When a \textsf{Tide} server is configured, all
output channels of a view stream in parallel to one window on the wall. In
\cite{deflect}, we have shown interactive framerates for a 24 megapixel
resolution over a WAN link. \textsf{Deflect} events are translated and injected
into the \textsf{Equalizer} event flow, allowing seamless application
integration.

@@

We investigated the scalability of individual decomposition modes by running the
same experiment using a varying number of render nodes (2-9), as well as one
dedicated application/display node. We subsequently summed up the duration of
all rendered frames for each run (\fig{fCompounds}). We provide the results of
the load balancer providing the better result for each application and compound.
The simpler \textsf{tree\_equalizer} outperforms in almost all cases the
load-grid-driven \textsf{load\_equalizer}, except for sort-first volume
rendering where the load in the region of interest is relatively uniform. This
counterintuitive result once more proves that simple algorithms often outperform
theoretically better, but more complex implementations.

@@

This figure gives an impression of how individual modes perform on heterogeneous
systems. In this case the tree equalizer performs best (\fig{fEqualizers}
(left)), as it allows us to a priori define how much {\em usage} it should make
of individual nodes, i.e., bias the allocation of rendering time in accordance
with the (simulated) compute resources. \fig{fEqualizers} (right), on the other
hand, illustrates how the tested decomposition modes perform on a system where
compute resources fluctuate randomly every frame, as can arguably be the case
for virutalized enviroments. For this scenario tile equalizer seems best suited,
as it performs load balancing implicitly and does not assume coherence of
available resources between frames. The simpler tree equalizer also outperforms
the load equalizer in this experiment.

@@

\begin{IEEEbiography}[{
\includegraphics[width=1in,height=1.25in,clip,keepaspectratio]{images/stefan}}]{
Stefan Eilemann} works towards large-scale visualization for Exascale
simulations; interactive integration of simulation, analysis and visualization
as well as flexible frameworks for data sharing and dynamic allocation of
heterogenous resources. He was the technical manager of the Visualization Team
in the Blue Brain Project, is the CEO and founder of Eyescale, and the lead
developer of the \textsf{Equalizer} parallel rendering framework. He received
his masters degree in Computer Science from EPFL in 2015, and his Engineering
Diploma in Computer Science in 1998. He is currently working towards a PhD in
Computer Science at the Visualization and MultiMedia Lab at the University of
Zurich. \end{IEEEbiography}
